Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-11 write-ups/PLDI/paper.tex
Cited a bunch of graph papers.

\paragraph{The needs of irregular data-parallel algorithms}
Acceleration of general-purpose computations on graphics processing
units (GPUs) has tended to focus on \emph{regular} data-parallel
algorithms, for which work to be processed can be evenly split between
the workgroups that execute a GPU kernel ahead of time.  However, many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  There is growing interest in accelerating irregular
data-parallel algorithms on GPUs, particularly algorithms that
traverse and manipulate linked data structures, with several recent notable
successes~\cite{...}.

@@

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{...}, CUDA~\cite{...}  and
HSA~\cite{...}---specify almost no guarantees regarding scheduling of
workgroups, and implementations of these programming models do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
If a kernel is executed with more workgroups than there are compute
units then at least two workgroups must share a single compute unit.
The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
execution model~\cite{...}.  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{...}.  It is
evidently a simple model for device vendors to implement efficiently.

@@

OpenCL specifically makes no guarantees about fair scheduling between
workgroups executing the same kernel, stating~\cite{...}: \emph{``A
  conforming implementation may choose to serialize the workgroups so
  a correct algorithm cannot assume that workgroups will execute in
  parallel.  There is no safe and portable way to synchronize across
  the independent execution of workgroups since once in the work-pool,
  they can execute in any order.''}  OpenCL is not alone in this
regard: CUDA similarly provides no guarantees~\cite{...}.

>>>>

\paragraph{The needs of irregular data-parallel algorithms}
Acceleration of general-purpose computations on graphics processing
units (GPUs) has tended to focus on \emph{regular} data-parallel
algorithms, for which work to be processed can be evenly split between
the workgroups that execute a GPU kernel ahead of time.  However, many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  The last decade has seen growing interest in accelerating irregular
data-parallel algorithms on GPUs, particularly algorithms that
traverse and manipulate linked data structures~\cite{owens-persistent,TPO10,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/oopsla/SorensenDBGR16}.  \ADComment{Tyler, can you add relevant workstealing papers if not here already?}

@@

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{...}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and implementations of these programming models do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
If a kernel is executed with more workgroups than there are compute
units then at least two workgroups must share a single compute unit.
The simplest way for a GPU driver to handle this is via an \emph{occupancy-bound}
execution model~\cite{...}.  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{...}.  It is
evidently a simple model for device vendors to implement efficiently.

@@

OpenCL specifically makes no guarantees about fair scheduling between
workgroups executing the same kernel, stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
  conforming implementation may choose to serialize the workgroups so
  a correct algorithm cannot assume that workgroups will execute in
  parallel.  There is no safe and portable way to synchronize across
  the independent execution of workgroups since once in the work-pool,
  they can execute in any order.''}  \ADComment{Tyler: please provide CUDA reference with page} OpenCL is not alone in this
regard: CUDA similarly provides no guarantees~\cite{...}.
